\pdftitle

### 2017-05-17 maxNt.tex
new material on projection

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-17T10:31:16+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

@@

By marginalization we obtain the probability for the state of the sample:
\begin{gather}
  \label{eq:marginal}
  \p(\yx = \yr \| \yH) = \binom{n}{n\yrs}^{-1} \p(\yxs=\yrs \| \yH),
  \\
  \shortintertext{with}
  \label{eq:subpop_average}
  \p(\yxs=\yrs \| \yH) = \sum_{N \yRf =0}^{N}
    \p(\yxs = \yrs \|\yXf=\yRf, \yH)\,
  \p(\yXf=\yRf \| \yH),
  \\[2\jot]
  \p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}.
\end{gather}

The latter conditional probability is a hypergeometric distribution,
typical of \enquote{drawing without replacement} problems. The
combinatorial proof of the formulae above is in fact the same as for this
class of problems
\cites[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.
This conditional probability acts a

@@

\begin{figure}[!t]
\centering
%\includegraphics[width=0.99\linewidth]{pop_sample_projection.pdf}\\
\includegraphics[width=0.99\linewidth]{pop_sample_projection2.pdf}%
\caption{hyperg}
\label{fig:hypergeom_proj}
\end{figure}%hypergeometric_identity.nb
% \end{wrapfigure}
Our initial knowledge (ignorance) $\yH$ has two important properties
derivable from the formulae above.

First, our initial expectations for the averages of activity products
\emph{are the same for the sample and for the full population}:
\begin{equation}
  \label{eq:expe_products}
\expe{\sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}} \| \yH}
=
\expe{\av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}} \| \yH}
%\\[2\jot]
%{}
=
\binom{n}{m}^{-1}
%\frac{(n-m)!}{n!}
\sum_{n \yrs=0}^n %m!\, 
\binom{n \yrs}{m}\, \p(\yxs=\yrs \| \yH)
% \sum_{N \yRf=0}^N %m!\, 
% \binom{N-m}{N \yRf-m}
% \binom{N}{N \yRf}^{-1} \p(\yXf=\yRf \| \yH)
=
\binom{N}{m}^{-1}
%\frac{(N-m)!}{N!}
\sum_{N \yRf=0}^N %m!\, 
\binom{N \yRf}{m}\, \p(\yXf=\yRf \| \yH).
\end{equation}
Raw moments like $\expe{\yxs^m}$, which can be written in terms of the
product expectations above via \eqn~\eqref{eq:expe_products}, similarly
satisfy $\expe{\yxs^m} \to \expe{\yXf^m}$ as $n \to N$.

Second, upon observation of a sample average, say $\yxxs=a$, the updated
expectations for such average in the population \emph{and in any new
  sample} will usually be shifted towards the observed value, as follows
from Bayes's theorem and the formulae above.

These two properties express the fact that the sample is
\emph{representative} of the population. Our initial knowledge $\yH$ thus
assumes that by observing a sample we can make predictions about the full
population and about new samples. Contrast this with an assumption that
assigns an independent probability to each neuron: in this case, the
observation of one sample does not change our initial probability about
another -- we can't learn from observation.

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-17T12:40:23+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\newcommand*{\ypp}{\varPi}

@@

By marginalization we obtain the probability for the state of the sample:
\begin{gather}
  \label{eq:marginal}
  \p(\yx = \yr \| \yH) = \binom{n}{n\yrs}^{-1} \p(\yxs=\yrs \| \yH),
  \\
  \shortintertext{with}
  \label{eq:subpop_average}
  \p(\yxs=\yrs \| \yH) = \sum_{N \yRf =0}^{N}
    \p(\yxs = \yrs \|\yXf=\yRf, \yH)\,
  \p(\yXf=\yRf \| \yH),
  \\[2\jot]
  \p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}
  \defs \ypp(\yrs\|\yRf).
\end{gather}

The latter conditional probability is a hypergeometric distribution,
typical of \enquote{drawing without replacement} problems. The
combinatorial proof of the formulae above is in fact the same as for this
class of problems
\cites[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.

The conditional probability $\ypp(\yrs,\yRf)$ connects the spaces of the
sample variable $\yXf$ and of the population variable $\yxs$. It is a
projector $\ypp$ of probabilities $p$ for $\yXf$ onto marginal
probabilities for $\yxs$:
\begin{equation}
  \label{eq:projection_P}
  \ypp\colon p(\yXf=\dotv) \to p_*(\yxs=\dotv)
  \defd \sum_{N\yRf=0}^N \ypp(\dotv\|\yRf) p(\yRf).
\end{equation}
It is also a pull-back $\ypp^*$ of functions $f$ of $\yxs$ to functions of
$\yXf$:
\begin{equation}
  \label{eq:pullback_P}
  \ypp^*\colon f(\yxs=\dotv) \to f^*(\yXf=\dotv) \defd \sum_{N\yrs=0}^n f(\yrs)\ypp(\yrs\|\dotv),
\end{equation}
with a caveat: if $f$ is a physical quantity and doesn't depend on our
knowledge, $f_*$ is \emph{not} a physical quantity but a conditional
expectation, $f^*(\yRf) \equiv \expe{f(\yxs)\|\yXf=\yRf,\yH}$, and depends
on our knowledge.

\begin{figure}[!t]
\centering
%\includegraphics[width=0.99\linewidth]{pop_sample_projection.pdf}\\
\includegraphics[width=0.99\linewidth]{pop_sample_projection2.pdf}%
\caption{Log-plot of the hypergeometric distribution
  $%\p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1} \defs
  \ypp(\yrs\|\yRf)$ for $N=5000$, $n=200$. (Band artifacts may appear in the
  colourbar depending on your \textsc{pdf} viewer.)}
\label{fig:hypergeom_proj}
\end{figure}%hypergeometric_identity.nb
A look at a plot of the hypergeometric distribution $\ypp(\yrs\|\yXf)$, for
example \fig~\ref{fig:hypergeom_proj}, reveals that the maps $\ypp$ and
$\ypp^*$ are \enquote{fuzzy identities} between the $\yXf$ space
$\set{0,\dotsc,N}$ and $\yxs$-space $\set{0,\dotsc,n}$. When $n=N$ they are
exact identities. We thus have very roughly that
\begin{equation}
  \label{eq:roughly_equal_nN}
  \p(\yxs=a) \sim \p(\yXf=a),\qquad
\expeb{f(\yxs)} \sim \expeb{f(\yXf)}.
\end{equation}
These formulae express the fact that \emph{the sample is representative of
  the population}, and vice versa. Our initial knowledge $\yH$, which fully
determines $\ypp$, thus assumes that by observing a sample we can make
predictions about the full population, and thence about new samples.
Upon observation of a sample average, say $\yxxs=a$, the updated
expectations for such average in the population and in any new sample will
usually be shifted towards the observed value, as follows from Bayes's
theorem and the formulae above.

Note that the approximate formulae~\eqref{eq:roughly_equal_nN} say more
than the limits $\p(\yxs=a) \to \p(\yXf=a)$,
$\expeb{f(\yxs)} \to \expeb{f(\yXf)}$ as $n\to N$. These limits are
trivially valid because as $n\to N$ the sample becomes the full population.
For conditional probabilities $\p(\yxs = \yrs \|\yXf=\yRf)$ very different
from the fuzzy identity $\ypp(\yrs\|\yRf)$ (see
\fig~\ref{fig:hypergeom_proj}) these limits would hold, but the
probabilities and expectations from a small sample could wildly differ from those
from the population.

\bigskip***

with the property that
\begin{equation}
  \label{eq:expectation_pullback}
  \expeb{f(\yxs) \|\yH} = \expeb{f^*(\yXf)\|\yH},
  \qquad\text{that is,}\qquad
  \sum_{n\yrs=0}^n f(\yrs)\,p_*(\yxs=\yrs) = \sum_{N\yRf=0}^N f^*(\yRf)\,p(\yXf=\yRf).
\end{equation}

A special case of the pull-back above is the function
$f(\yxs) \defd \sav{\yx\dots\yx}=\binom{\yxs}{m}/\binom{n}{m}$, for which
\begin{equation}
  \label{eq:pullback_m_expectations}
  \ypp^*\colon \sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}}
  \mapsto \bigl( \sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}}\bigr)^*
  \equiv
  \av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}};
\end{equation}
thus, the averages of activity products \emph{are the same for the sample
  and for the full population}:
\begin{equation}
  \label{eq:expe_products}
\expe{\sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}} \| \yH}
=
\expe{\av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}} \| \yH}
%\\[2\jot]
%{}
=
\binom{n}{m}^{-1}
%\frac{(n-m)!}{n!}
\sum_{n \yrs=0}^n %m!\, 
\binom{n \yrs}{m}\, \p(\yxs=\yrs \| \yH)
% \sum_{N \yRf=0}^N %m!\, 
% \binom{N-m}{N \yRf-m}
% \binom{N}{N \yRf}^{-1} \p(\yXf=\yRf \| \yH)
=
\binom{N}{m}^{-1}
%\frac{(N-m)!}{N!}
\sum_{N \yRf=0}^N %m!\, 
\binom{N \yRf}{m}\, \p(\yXf=\yRf \| \yH).
\end{equation}

This is true, for example, for the raw moments $\expe{\yxs^m}$ and
$\expe{\yXf^m}$, which can be written in terms of the product expectations
via \eqn~\eqref{eq:expe_products}.

@@
